---
title: "PNAS Nexus paper"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Let's test this out

```{r}
library(tidyverse)
library(magrittr)
```

Let's load the file
```{r}
myTable = read_tsv("cytokine_correlation.txt")
head(myTable)
```

```{r}
dim(myTable)
```

OK so we have 130/2=65 parameters to compare, which is correct.

Let's split them into 2 tables, one for case and one for control

```{r}
myTable$GROUP
myCase = myTable[which(myTable$GROUP == "CASE"),] %>% as.data.frame(.)
myControl = myTable[which(myTable$GROUP == "CONTROL"),] %>% as.data.frame(.)
```

OK, now let's loop through the entire tables and do our correlations based on them

```{r}
myvariables = {}

for (x in seq(from = 2, to = dim(myControl)[2],by=2)){
     case_blood = myCase[,x] %>% as.data.frame(.)
  control_blood = myControl[,x] %>% as.data.frame(.)
      case_saliva = myCase[,(x+1)] %>% as.data.frame(.)
   control_saliva = myControl[,(x+1)] %>% as.data.frame(.)
   
   spearman_case = cor.test(as.numeric(case_blood[,1]),as.numeric(case_saliva[,1]), method = c("spearman"),exact = FALSE)
   spearman_case = spearman_case$p.value
   spearman_control = cor.test(as.numeric(control_blood[,1]),as.numeric(control_saliva[,1]), method = c("spearman"),exact = FALSE)
   spearman_control = spearman_control$p.value
   
   if (spearman_case < 0.05 || spearman_control < 0.05) {
     myvariables = c(myvariables,x)
   }
}

for (x in myvariables){
currentCase = data.frame(myCase[,x],myCase[,(x+1)]) %$% cor.test(as.numeric(.[,1]),as.numeric(.[,2]), method = c("spearman"),exact = FALSE)
currentControl = data.frame(myControl[,x],myControl[,(x+1)]) %$% cor.test(as.numeric(.[,1]),as.numeric(.[,2]), method = c("spearman"),exact = FALSE)
print(paste0("Rho correlation in 'case' between ",
             colnames(myCase)[x]," and ",
             colnames(myCase)[x+1]," is ",
             currentCase$estimate, " and the pvalue is ",currentCase$p.value))

print(paste0("Rho correlation in 'control' between ",
             colnames(myControl)[x]," and ",
             colnames(myControl)[x+1]," is ",
             currentControl$estimate, " and the pvalue is ",currentControl$p.value))
}

```
Some variable clean up
```{r}
rm(control_blood,control_saliva,currentCase,currentControl,myvariables,spearman_case,spearman_control,x,case_blood,case_saliva,myCase,myControl)
```



Let's get the cytokines correlation
```{r}
ArrangedCytokine = read_tsv(file = "./Kuwait_project_blood_saliva.txt") %>% as.data.frame(.) %>% .[,which(colnames(.) %in% c("#SampleID","group2","APRIL_blood","APRIL_saliva","BAFF_blood","BAFF_saliva","BLCCXCL13_blood","BLCCXCL13_saliva","CD30_blood","CD30_saliva","CD40L_blood","CD40L_saliva","ENA78CXCL5_blood","ENA78CXCL5_saliva","Eotaxin2CCL24_blood","Eotaxin2CCL24_saliva","Eotaxin3CCL26_blood","Eotaxin3CCL26_saliva","EotaxinCCL11_blood","EotaxinCCL11_saliva","FGF2_blood","FGF2_saliva","FractalkineCX3CL1_blood","FractalkineCX3CL1_saliva","GCSFCSF3_blood","GCSFCSF3_saliva","GMCSF_blood","GMCSF_saliva","GROalphaCXCL1_blood","GROalphaCXCL1_saliva","HGF_blood","HGF_saliva","IFNalpha_blood","IFNalpha_saliva","IFNgamma_blood","IFNgamma_saliva","IL1alpha_blood","IL1alpha_saliva","IL1beta_blood","IL1beta_saliva","IL2_blood","IL2_saliva","IL2R_blood","IL2R_saliva","IL3_blood","IL3_saliva","IL4_blood","IL4_saliva","IL5_blood","IL5_saliva","IL6_blood","IL6_saliva","IL7_blood","IL7_saliva","IL8CXCL8_blood","IL8CXCL8_saliva","IL9_blood","IL9_saliva","IL10_blood","IL10_saliva","IL12p70_blood","IL12p70_saliva","IL13_blood","IL13_saliva","IL15_blood","IL15_saliva","IL16_blood","IL16_saliva","IL17ACTLA8_blood","IL17ACTLA8_saliva","IL18_blood","IL18_saliva","IL20_blood","IL20_saliva","IL21_blood","IL21_saliva","IL22_blood","IL22_saliva","IL23_blood","IL23_saliva","IL27_blood","IL27_saliva","IL31_blood","IL31_saliva","IP10CXCL10_blood","IP10CXCL10_saliva","ITACCXCL1_blood","ITACCXCL11_saliva","LIF_blood","LIF_saliva","MCP1CCL2_blood","MCP1CCL2_saliva","MCP2CCL8_blood","MCP2CCL8_saliva","MCP3CCL7_blood","MCP3CCL7_saliva","MCSF_blood","MCSF_saliva","MDC_blood","MDC_saliva","MIF_blood","MIF_saliva","MIGCXCL9_blood","MIGCXCL9_saliva","MIP1alphaCCL3_blood","MIP1alphaCCL3_saliva","MIP1betaCCL4_blood","MIP1betaCCL4_saliva","MIP3alphaCCL20_blood","MIP3alphaCCL20_saliva","MMP1_blood","MMP1_saliva","NGFbeta_blood","NGFbeta_saliva","SCF_blood","SCF_saliva","SDF1alpha_blood","SDF1alpha_saliva","TNFalpHa_blood","TNFalpha_saliva","TNFbeta_blood","TNFbeta_saliva","TNFRII_blood","TNFRII_saliva","TRAIL_blood","TRAIL_saliva","TSLP_blood","TSLP_saliva","TWEAK_blood","TWEAK_saliva","VEGFA_blood","VEGFA_saliva"))] %>% as.data.frame(.) # Each cytokine is arranged in this list. Starts with blood, then saliva

bloodCytokines = ArrangedCytokine %>% .[,which(grepl("_blood", colnames(ArrangedCytokine), fixed=TRUE))] %>% cbind(ArrangedCytokine$`#SampleID`,ArrangedCytokine$group2,.)
colnames(bloodCytokines)[1] = "SampleID"
colnames(bloodCytokines)[2] = "group2"

salivaCytokines = ArrangedCytokine %>% .[,which(grepl("_saliva", colnames(ArrangedCytokine), fixed=TRUE))] %>% cbind(ArrangedCytokine$`#SampleID`,ArrangedCytokine$group2,.)
colnames(salivaCytokines)[1] = "SampleID"
colnames(salivaCytokines)[2] = "group2"

# Remove the designation of either blood or saliva, to make it easier to plot
colnames(bloodCytokines) = gsub('_blood', '', colnames(bloodCytokines))
colnames(salivaCytokines) = gsub('_saliva', '', colnames(bloodCytokines))

colnames(salivaCytokines) == colnames(bloodCytokines) # TRUE, so all columns are the same and are arranged the same
salivaCytokines$SampleID == bloodCytokines$SampleID  # TRUE, so all rows are the same and are arranged the same

```

Great so it is nicely arranged now. Let's create another table, and insert the pvalues of the correlations in it

```{r}
cytokineList = colnames(salivaCytokines)[3:dim(salivaCytokines)[2]]
mySpearman = data.frame(matrix(NA, nrow = length(cytokineList), ncol = 3))
mySpearman[,1] = cytokineList

for (x in 1:dim(mySpearman)[1]){
  currentCytokine = mySpearman[x,1]
  bloodCytokine = bloodCytokines %>% .[,which(colnames(.) %in% currentCytokine)]
  salivaCytokine = salivaCytokines %>% .[,which(colnames(.) %in% currentCytokine)]
  
  currentSpearman = cor.test(bloodCytokine,salivaCytokine,method="spearman",exact=FALSE)
  mySpearman[x,2] = currentSpearman$estimate
  mySpearman[x,3] = currentSpearman$p.value
}

rm(bloodCytokine,currentCytokine,x,salivaCytokine,ArrangedCytokine,myTable) ## Variable clean up
```

So we want to create a heatmap for the statistically sig cytokines in saliva. We are trying to show upregulation and downregulation. So first run a kruskall wallis (Mann Whitney) on Yes/No COVID. Those that are stat sig, per cytokine, take the values, turn them to z-scores, and put them in a heatmap.

Let's see what is sig and what is not in saliva and blood

```{r message=FALSE, warning=FALSE}

SalivaZscoreTable = data.frame(matrix(NA, nrow = length(cytokineList), ncol = dim(salivaCytokines)[1]+1))
rownames(SalivaZscoreTable) = cytokineList
colnames(SalivaZscoreTable) = c(salivaCytokines$SampleID,"pvalue")

salivaDunn = data.frame(matrix(NA, nrow = 0, ncol = 13))
colnames(salivaDunn) = c(".y.","group1","group2","n1","n2","statistic","p","p.adj","p.adj.signif","y.position","groups","xmin","xmax")

for (x in 1:length(cytokineList)){
currentCytokine = salivaCytokines[,cytokineList[x]]
categories = salivaCytokines$group2
kstest = kruskal.test(currentCytokine,categories)

#salivaDunn = rbind(salivaDunn,currentDunn)
currentCytokine = currentCytokine+(abs(min(currentCytokine,na.rm=TRUE)) + 1) #Take all the cytokines, and add to them the absolute value of the minimum value, plus 1
currentCytokine = log10(currentCytokine) #log10

#kstest = aov(currentCytokine~categories) %>% summary(.) %>% .[[1]] %>% .$`Pr(>F)` %>% .[1]
#kstest$p.value = kstest

zscoreTable = scale(currentCytokine) #Shift all the values by min+1, then convert to z-scores. This will change all the sco
SalivaZscoreTable[x,] = c(zscoreTable,kstest$p.value)

currentDunn = rstatix::t_test(data.frame(zscoreTable,categories),
                                 formula=zscoreTable~categories,p.adjust.method="BH") %>% 
  rstatix::add_xy_position(x="categories")
currentDunn$.y. = cytokineList[x]

salivaDunn = rbind(salivaDunn,currentDunn)
}

# Keep only the significant ones
#SalivaZscoreTable = SalivaZscoreTable[which(SalivaZscoreTable$pvalue < 0.05),] # now you have a table with only the signficantly different between the 3 categories, then transformed into z-scores

sigSalivaCytokines = rownames(SalivaZscoreTable)


salivaDunnSigCytokines = salivaDunn %>% .$.y. %>% unique(.)
salivaDunnSigOnly = salivaDunn %>% .[which(.$.y. %in% salivaDunnSigCytokines),] %>% .[which(.$p.adj <0.05),]

SalivaZscoreDunnSigOnly = SalivaZscoreTable %>% .[rownames(.) %in% salivaDunnSigCytokines,] %>% rownames_to_column(var = "FeatureID") %>%
  pivot_longer(cols = -FeatureID, names_to = "SampleID", values_to = "Value") 

SalivaZscoreDunnSigOnly = left_join(SalivaZscoreDunnSigOnly,salivaCytokines[,1:2])
SalivaZscoreDunnSigOnly$SampleID = NULL
SalivaZscoreDunnSigOnly = SalivaZscoreDunnSigOnly %>% .[-which(is.na(.$group2)),]

SalivaZscoreDunnSigOnly = SalivaZscoreDunnSigOnly %>% .[which( .$FeatureID %in% c("CD40L", "IP10CXCL10") ),]


ggplot(SalivaZscoreDunnSigOnly, aes(x = FeatureID, y = Value, fill = group2)) + geom_boxplot() + theme_bw()  #ggpubr::stat_pvalue_manual(salivaDunnSigOnly, label = "p.adj.signif", y.position = "y.position",tip.length = 0.01)


rm(x,zscoreTable)
```

```{r}

bloodZscoreTable = data.frame(matrix(NA, nrow = length(cytokineList), ncol = dim(bloodCytokines)[1]+1))
rownames(bloodZscoreTable) = cytokineList
colnames(bloodZscoreTable) = c(bloodCytokines$SampleID,"pvalue")

for (x in 1:length(cytokineList)){
currentCytokine = bloodCytokines[,cytokineList[x]]
categories = bloodCytokines$group2
kstest = kruskal.test(currentCytokine,categories)
currentCytokine = currentCytokine+(abs(min(currentCytokine,na.rm=TRUE)) + 1) #Take all the cytokines, and add to them the absolute value of the minimum value, plus 1
currentCytokine = log10(currentCytokine) #log10
zscoreTable = scale(currentCytokine) #convert to z-scores
bloodZscoreTable[x,] = c(zscoreTable,kstest$p.value)
}

# Keep only the significant ones
#bloodZscoreTable = bloodZscoreTable[which(bloodZscoreTable$pvalue < 0.05),] # now you have a table with only the signficantly different between the 3 categories, then transformed into z-scores

sigBloodCytokines = rownames(bloodZscoreTable)

BloodZscoreForPlotting = bloodZscoreTable %>% .[rownames(.) %in% c("CD40L", "IP10CXCL10"),] %>% rownames_to_column(var = "FeatureID") %>%
  pivot_longer(cols = -FeatureID, names_to = "SampleID", values_to = "Value") 

BloodZscoreForPlotting = BloodZscoreForPlotting %>% .[-which(.$SampleID %in% "pvalue"),]

BloodZscoreForPlotting = left_join(BloodZscoreForPlotting,bloodCytokines[,1:2])
BloodZscoreForPlotting$SampleID = NULL
BloodZscoreForPlotting = BloodZscoreForPlotting[-60,]

ggplot(BloodZscoreForPlotting, aes(x = FeatureID, y = Value, fill = group2)) + geom_boxplot() + theme_bw() 

rm(x,kstest)
```

Let's draw heatmaps



Let's work on the graphing now. Will need a complex heatmap. The generated heatmap in this notebook is based off the tutorial by: https://github.com/kevinblighe/E-MTAB-6141

```{r}
 if (!requireNamespace('ComplexHeatmap', quietly = TRUE))
    BiocManager::install('ComplexHeatmap')
library('ComplexHeatmap')
```

Let's start with saliva

```{r}
# Need to format 2 objects. One for the scores, where the samplesa re in colums and the cytokines in rows
mat = SalivaZscoreTable
mat$pvalue = NULL

# Other object is the one with the metadata. Sample IDs in rows, and categories as columns
metadata = salivaCytokines
rownames(metadata) = salivaCytokines$SampleID
metadata$SampleID = NULL
```

Preparation for the heatmap
```{r}
# Create the colors, and their breaks
myCol = colorRampPalette(c('dodgerblue', 'black', 'yellow'))(100) 
myBreaks = seq(-3, 3, length.out = 100)

#Let's create the legend annotation
ann = data.frame(
    Severity = metadata$group2,
    stringsAsFactors = FALSE)

colours = list(
    severity = c('Control' = 'pink', 'mild_moderate' = 'pink4', 'Severe' = 'red'))
```

Let's see if there is a way to do hierarchical clustering and then join them all together
```{r}
healthy = metadata[which(metadata$group2 == "Control"),] %>% rownames(.)
healthy2 = mat %>% t(.) %>% as.data.frame(.) %>% .[which(rownames(.) %in% healthy),]
healthy$group2 = NULL
healthy = dist(healthy2) %>% hclust(.) %>% .$labels

mild_moderate = metadata[which(metadata$group2 == "mild_moderate"),]
mild_moderate$group2 = NULL
mild_moderate = dist(mild_moderate) %>% hclust(.) %>% .$labels

severe = metadata[which(metadata$group2 == "Severe"),]
severe$group2 = NULL
severe = dist(severe) %>% hclust(.) %>% .$labels
```


```{r}
severity = metadata$group2 %>% as.data.frame(.)
colnames(severity) = "severity"
severity$severity = factor(severity$severity,
    levels = c('Control', 'mild_moderate', 'Severe'))
rownames(severity) = rownames(metadata)

colAnn = HeatmapAnnotation(
    df = severity,
    which = 'col', # 'col' (samples) or 'row' (gene) annotation?
    na_col = 'white', # default colour for any NA values in the annotation data-frame, 'ann'
    col = colours,
    annotation_height = 0.6,
    annotation_width = unit(1, 'cm'),
    gap = unit(1, 'mm'),
    annotation_legend_param = list(
      severity = list(
        nrow = 4, # number of rows across which the legend will be arranged
        title = 'Severity',
        title_position = 'topleft',
        legend_direction = 'horizontal',
        title_gp = gpar(fontsize = 8, fontface = 'bold'),
        labels_gp = gpar(fontsize = 8, fontface = 'bold'))))


hmap = Heatmap(as.matrix(mat),
               name = "Cytokine\nz-score",
               col=circlize::colorRamp2(myBreaks,myCol),
               heatmap_legend_param = list(
                  color_bar = 'continuous',
                  legend_direction = 'vertical',
                  legend_width = unit(8, 'cm'),
                  legend_height = unit(5.0, 'cm'),
                  title_position = 'topcenter',
                  title_gp=gpar(fontsize = 8, fontface = 'bold'),
                  labels_gp=gpar(fontsize = 8, fontface = 'bold')),
                  
               show_column_names = FALSE,
               cluster_columns = FALSE,
               cluster_column_slices = TRUE,
               show_column_dend = FALSE,
               column_order = c(healthy,mild_moderate,severe),
               
               
              top_annotation = colAnn
               
               )
salivaryHeatMap = draw(hmap, column_title = "Salivary cytokines")
salivaryHeatMap
```


```{r}
rm(ann,categories,colAnn,colours,currentCytokine,healthy,hmap,mat,mild_moderate,myBreaks,myCol,SalivaZscoreTable,severe,severity,zscoreTable)
```

Let's do the blood one now.


```{r}
# Need to format 2 objects. One for the scores, where the samplesa re in colums and the cytokines in rows
mat = bloodZscoreTable
mat$pvalue = NULL

# Other object is the one with the metadata. Sample IDs in rows, and categories as columns
metadata = bloodCytokines
rownames(metadata) = bloodCytokines$SampleID
metadata$SampleID = NULL
```

Preparation for the heatmap
```{r}
# Create the colors, and their breaks
myCol = colorRampPalette(c('dodgerblue', 'black', 'yellow'))(100) 
myBreaks = seq(-3, 3, length.out = 100)

#Let's create the legend annotation
ann = data.frame(
    severity = metadata$group2,
    stringsAsFactors = FALSE)

colours = list(
    severity = c('Control' = 'pink', 'mild_moderate' = 'pink4', 'Severe' = 'red'))
```

Let's see if there is a way to do hierarchical clustering and then join them all together
```{r}
healthy = metadata[which(metadata$group2 == "Control"),]
healthy$group2 = NULL
healthy = dist(healthy) %>% hclust(.) %>% .$labels

mild_moderate = metadata[which(metadata$group2 == "mild_moderate"),]
mild_moderate$group2 = NULL
mild_moderate = dist(mild_moderate) %>% hclust(.) %>% .$labels

severe = metadata[which(metadata$group2 == "Severe"),]
severe$group2 = NULL
severe = dist(severe) %>% hclust(.) %>% .$labels
```


```{r}
severity = metadata$group2 %>% as.data.frame(.)
colnames(severity) = "severity"
severity$severity = factor(severity$severity,
    levels = c('Control', 'mild_moderate', 'Severe'))
rownames(severity) = rownames(metadata)

colAnn = HeatmapAnnotation(
    df = severity,
    which = 'col', # 'col' (samples) or 'row' (gene) annotation?
    na_col = 'white', # default colour for any NA values in the annotation data-frame, 'ann'
    col = colours,
    annotation_height = 0.6,
    annotation_width = unit(1, 'cm'),
    gap = unit(1, 'mm'),
    annotation_legend_param = list(
      severity = list(
        nrow = 4, # number of rows across which the legend will be arranged
        title = 'Severity',
        title_position = 'topleft',
        legend_direction = 'vertical',
        title_gp = gpar(fontsize = 8, fontface = 'bold'),
        labels_gp = gpar(fontsize = 8, fontface = 'bold'))))


hmap = Heatmap(as.matrix(mat),
               name = "Cytokine\nz-score",
               col=circlize::colorRamp2(myBreaks,myCol),
               heatmap_legend_param = list(
                  color_bar = 'continuous',
                  legend_direction = 'vertical',
                  legend_width = unit(8, 'cm'),
                  legend_height = unit(5.0, 'cm'),
                  title_position = 'topcenter',
                  title_gp=gpar(fontsize = 8, fontface = 'bold'),
                  labels_gp=gpar(fontsize = 8, fontface = 'bold')),
                  
               show_column_names = FALSE,
               cluster_columns = FALSE,
               cluster_column_slices = TRUE,
               show_column_dend = FALSE,
               column_order = c(healthy,mild_moderate,severe),
               
               
              top_annotation = colAnn
               
               )
draw(hmap, column_title = "Blood cytokines")
```
```{r}
rm(severe,myCol,myBreaks,mild_moderate,metadata,mat,healthy,severity,ann,bloodZscoreTable,colAnn,colours,hmap)
```

```{r}

# Let's start by filtering out sig salivary cytokines to only those that are sig
sigSalivaCytokines2 = salivaCytokines %>% .[,which(colnames(.) %in% c("group2",sigSalivaCytokines))] %>% pivot_longer(.,cols = !group2,names_to = "Names",values_to = "Value")
sigBloodCytokines2  = bloodCytokines %>% .[,which(colnames(.) %in% c("group2",sigBloodCytokines))] %>% pivot_longer(.,cols = !group2,names_to = "Names",values_to = "Value")
```

Lets start with saliva
```{r}
log10List = data.frame(matrix(NA, nrow = 0, ncol = 5))
colnames(log10List) = c("group2","Names","Value","Value2","Value3")

for (x in (sigSalivaCytokines2$Names %>% unique(.))){
currentList = sigSalivaCytokines2 %>% .[.$Names == x,]
theValues = currentList$Value
theminimum = min(theValues) %>% abs(.)+1
currentList$Value2 = theValues+theminimum
currentList$Value3 = log10(currentList$Value2)
log10List = rbind(log10List,currentList)
}

p2 = ggplot(log10List, aes(x=Names, y=Value3, fill=group2)) + 
    geom_boxplot() +
    facet_wrap(~Names, scale="free") + theme_minimal()
p2
```

Lets do blood
```{r}
log10List = data.frame(matrix(NA, nrow = 0, ncol = 5))
colnames(log10List) = c("group2","Names","Value","Value2","Value3")

for (x in (sigBloodCytokines2$Names %>% unique(.))){
currentList = sigBloodCytokines2 %>% .[.$Names == x,]
theValues = currentList$Value
theminimum = min(theValues,na.rm=TRUE) %>% abs(.)+1
currentList$Value2 = theValues+theminimum
currentList$Value3 = log10(currentList$Value2)
log10List = rbind(log10List,currentList)
}

colnames(log10List)[2] = "Cytokines"

library(rstatix)
stat.test = log10List %>%
  group_by(Cytokines) %>%
  wilcox_test(Value3 ~ group2) %>%
  adjust_pvalue(method = "fdr") %>%
  add_significance("p.adj") %>% add_xy_position(fun = "median", x = "Cytokines")

p2 = ggplot(log10List, aes(x=Cytokines, y=Value3, fill=group2)) + geom_boxplot() + facet_wrap(~Cytokines, scale="free") + theme_minimal() + ggpubr::stat_pvalue_manual(stat.test, label = "p.adj.signif",hide.ns = TRUE)
p2
```

Next, let's do correlograms

variable clean up
```{r}
rm(list=ls())
```


```{r}
library("corrplot")
library("tidyverse")
library("RColorBrewer")
```

Graphing

```{r}
cytokines = read_tsv("./Kuwait_project_for_correlation 2.txt",show_col_types = FALSE) %>% as.data.frame(.)
rownames(cytokines) = cytokines$`#SampleID`
cytokines$`#SampleID` = NULL

bloodCytokines = c("group2","covid_nocovid","APRIL_blood","BAFF_blood","BLCCXCL13_blood","CD30_blood","CD40L_blood","ENA78CXCL5_blood","Eotaxin2CCL24_blood","Eotaxin3CCL26_blood","EotaxinCCL11_blood","FGF2_blood","FractalkineCX3CL1_blood","GCSFCSF3_blood","GMCSF_blood","GROalphaCXCL1_blood","HGF_blood","IFNalpha_blood","IFNgamma_blood","IL10_blood","IL12p70_blood","IL13_blood","IL15_blood","IL16_blood","IL17ACTLA8_blood","IL18_blood","IL1alpha_blood","IL1beta_blood","IL2_blood","IL20_blood","IL21_blood","IL22_blood","IL23_blood","IL27_blood","IL2R_blood","IL3_blood","IL31_blood","IL4_blood","IL5_blood","IL6_blood","IL7_blood","IL8CXCL8_blood","IL9_blood","IP10CXCL10_blood","ITACCXCL11_blood","LIF_blood","MCP1CCL2_blood","MCP2CCL8_blood","MCP3CCL7_blood","MCSF_blood","MDC_blood","MIF_blood","MIGCXCL9_blood","MIP1alphaCCL3_blood","MIP1betaCCL4_blood","MIP3alphaCCL20_blood","MMP1_blood","NGFbeta_blood","SCF_blood","SDF1alpha_blood","TNFalpha_blood","TNFbeta_blood","TNFRII_blood","TRAIL_blood","TSLP_blood","TWEAK_blood","VEGFA_blood")
bloodCytokines = cytokines[,which(colnames(cytokines) %in% bloodCytokines)]


salivaCytokines = c("group2","covid_nocovid","APRIL_salivary","BAFF_salivary","BLCCXCL13_salivary","CD30_salivary","CD40L_salivary","ENA78CXCL5_salivary","Eotaxin2CCL24_salivary","Eotaxin3CCL26_salivary","EotaxinCCL11_salivary","FGF2_salivary","FractalkineCX3CL1_salivary","GCSFCSF3_salivary","GMCSF_salivary","GROalphaCXCL1_salivary","HGF_salivary","IFNalpha_salivary","IFNgamma_salivary","IL10_salivary","IL12p70_salivary","IL13_salivary","IL15_salivary","IL16_salivary","IL17ACTLA8_salivary","IL18_salivary","IL1alpha_salivary","IL1beta_salivary","IL2_salivary","IL20_salivary","IL21_salivary","IL22_salivary","IL23_salivary","IL27_salivary","IL2R_salivary","IL3_salivary","IL31_salivary","IL4_salivary","IL5_salivary","IL6_salivary","IL7_salivary","IL8CXCL8_salivary","IL9_salivary","IP10CXCL10_salivary","ITACCXCL11_salivary","LIF_salivary","MCP1CCL2_salivary","MCP2CCL8_salivary","MCP3CCL7_salivary","MCSF_salivary","MDC_salivary","MIF_salivary","MIGCXCL9_salivary","MIP1alphaCCL3_salivary","MIP1betaCCL4_salivary","MIP3alphaCCL20_salivary","MMP1_salivary","NGFbeta_salivary","SCF_salivary","SDF1alpha_salivary","TNFalpha_salivary","TNFbeta_salivary","TNFRII_salivary","TRAIL_salivary","TSLP_salivary","TWEAK_salivary","VEGFA_salivary")
salivaCytokines = cytokines[,which(colnames(cytokines) %in% salivaCytokines)]

cytokines = c("APRIL","BAFF","BLCCXCL13","CD30","CD40L","ENA78CXCL5","Eotaxin2CCL24","Eotaxin3CCL26","EotaxinCCL11","FGF2","FractalkineCX3CL1","GCSFCSF3","GMCSF","GROalphaCXCL1","HGF","IFNalpha","IFNgamma","IL10","IL12p70","IL13","IL15","IL16","IL17ACTLA8","IL18","IL1alpha","IL1beta","IL2","IL20","IL21","IL22","IL23","IL27","IL2R","IL3","IL31","IL4","IL5","IL6","IL7","IL8CXCL8","IL9","IP10CXCL10","ITACCXCL11","LIF","MCP1CCL2","MCP2CCL8","MCP3CCL7","MCSF","MDC","MIF","MIGCXCL9","MIP1alphaCCL3","MIP1betaCCL4","MIP3alphaCCL20","MMP1","NGFbeta","SCF","SDF1alpha","TNFalpha","TNFbeta","TNFRII","TRAIL","TSLP","TWEAK","VEGFA")
```

Let's create our correlation matrix

```{r}
correlationMatrix = matrix(nrow=65,ncol=65) %>% as.data.frame(.)
colnames(correlationMatrix) = cytokines
rownames(correlationMatrix) = cytokines

pvalueMatrix = correlationMatrix



# the columns are the saliva, rows are blood
```

```{r}
correlograph = function(bloodCytokines,salivaCytokines,mytitle){

for (currentBloodCytokine  in 1:65){
for (currentSalivaCytokine in 1:65){
  
currentBloodCytokineNums  = bloodCytokines[,currentBloodCytokine]
currentSalivaCytokineNums = salivaCytokines[,currentSalivaCytokine]

currentCorr = cor.test(currentBloodCytokineNums,currentSalivaCytokineNums,method="spearman",exact=FALSE)

correlationMatrix[currentBloodCytokine,    currentSalivaCytokine]     = currentCorr$estimate
     pvalueMatrix[currentBloodCytokine,    currentSalivaCytokine]     = currentCorr$p.value
  
}}
correlationGraph = corrplot(as.matrix(correlationMatrix),method="color",p.mat=as.matrix(pvalueMatrix),order="alphabet",sig.level = 0.05, insig="pch",pch.cex=.1,tl.col="black",tl.cex=0.5,title=mytitle)
  return(correlationGraph)}

correlographSigOnly = function(bloodCytokines,salivaCytokines,mytitle){

for (currentBloodCytokine  in 1:65){
for (currentSalivaCytokine in 1:65){
  
currentBloodCytokineNums  = bloodCytokines[,currentBloodCytokine]
currentSalivaCytokineNums = salivaCytokines[,currentSalivaCytokine]

currentCorr = cor.test(currentBloodCytokineNums,currentSalivaCytokineNums,method="spearman",exact=FALSE)

correlationMatrix[currentBloodCytokine,    currentSalivaCytokine]     = currentCorr$estimate
     pvalueMatrix[currentBloodCytokine,    currentSalivaCytokine]     = currentCorr$p.value
  
}}
correlationGraph = corrplot(as.matrix(correlationMatrix),method="color",p.mat=as.matrix(pvalueMatrix),order="alphabet",sig.level = 0.05, insig="blank",pch.cex=.1,tl.col="black",tl.cex=0.5,title=mytitle)
  return(correlationGraph)}
```

Let's graph COVID/No COVID

```{r}
COVIDbloodCytokines  = bloodCytokines  %>% .[which(.$covid_nocovid == "COVID"),]
COVIDbloodCytokines$group2 = NULL
COVIDbloodCytokines$covid_nocovid = NULL

COVIDsalivaCytokines = salivaCytokines %>% .[which(.$covid_nocovid == "COVID"),]
COVIDsalivaCytokines$group2 = NULL
COVIDsalivaCytokines$covid_nocovid = NULL

noCOVIDbloodCytokines  = bloodCytokines  %>% .[which(.$covid_nocovid == "NoCOVID"),]
noCOVIDbloodCytokines$group2 = NULL
noCOVIDbloodCytokines$covid_nocovid = NULL

noCOVIDsalivaCytokines = salivaCytokines %>% .[which(.$covid_nocovid == "NoCOVID"),]
noCOVIDsalivaCytokines$group2 = NULL
noCOVIDsalivaCytokines$covid_nocovid = NULL
```

```{r echo=TRUE, message=FALSE}
correlograph(noCOVIDbloodCytokines,noCOVIDsalivaCytokines,"Controls")
```

```{r}
correlograph(COVIDbloodCytokines,COVIDsalivaCytokines,"COVID")
```

```{r}
correlographSigOnly(noCOVIDbloodCytokines,noCOVIDsalivaCytokines,"Controls")
```

```{r}
correlographSigOnly(COVIDbloodCytokines,COVIDsalivaCytokines,"COVID")
```

Let's graph COVID/No COVID

```{r}
MMbloodCytokines  = bloodCytokines  %>% .[which(.$group2 == "mild_moderate"),]
MMbloodCytokines$group2 = NULL
MMbloodCytokines$covid_nocovid = NULL

MMsalivaCytokines = salivaCytokines %>% .[which(.$group2 == "mild_moderate"),]
MMsalivaCytokines$group2 = NULL
MMsalivaCytokines$covid_nocovid = NULL

severebloodCytokines  = bloodCytokines  %>% .[which(.$group2 == "Severe"),]
severebloodCytokines$group2 = NULL
severebloodCytokines$covid_nocovid = NULL

severesalivaCytokines = salivaCytokines %>% .[which(.$group2 == "Severe"),]
severesalivaCytokines$group2 = NULL
severesalivaCytokines$covid_nocovid = NULL
```


```{r}
correlograph(MMbloodCytokines,MMsalivaCytokines,"Mild/moderate symptoms")
```

```{r}
correlograph(severebloodCytokines,severesalivaCytokines,"Severe symptoms")
```

```{r}
correlographSigOnly(MMbloodCytokines,MMsalivaCytokines,"Mild/moderate symptoms")
```

```{r}
correlographSigOnly(severebloodCytokines,severesalivaCytokines,"Severe symptoms")
```

```{r}
correlographSigOnly(noCOVIDbloodCytokines,noCOVIDsalivaCytokines,"No COVID")
```
